Analysing Constraint Grammars with a SAT-solver

### 2016-02-10 paper/lrec/eval_futurework.tex
more stuff to paper
add stuff about new features

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
                      & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
\# rules      & 59              & 279      & 1185     \\ \hline
\# readings      & 276             & 2191     & 2012    \\ \hline
\# conflicts       & 1               & 1        & 51    \\ \hline
\clock{} all rules       & 7s              & 3min 22s    & 3h 26min    \\ \hline
\clock{} last rule       & 4s              & 7s                & 3min 40s    \\ \hline
% This is with just 2000ish variables for Finnish, below is with 5700

@@

\end{tabular}
\caption{Detecting rules that cannot be applied}
\label{table:res}
\end{table}

@@

The Spanish grammar had an erroneous set definition, where a word was
required to have almost all POS tags at the same time. After fixing
the set definition, our test did not reveal other problems.
The Dutch grammar had the following rules in the given order, making the latter impossible to apply:
\begin{itemize}
\item[] 
\begin{verbatim}REMOVE Adv IF (1 Noun) ;
REMOVE Adv IF (-1 Det) (0 Adj) (1 Noun) ;
\end{verbatim}
\end{itemize}

@@

% As another line of investigation, we want to detect more kinds of conflicts.
% For instance, say we have the following rules:

% \begin{itemize}
% \item[] \begin{verbatim}
% REMOVE verb IF (-1 det) ;
% REMOVE noun IF (-1 det) ;
% \end{verbatim}
% \end{itemize}

% With our symbolic sentence, these rules will be no problem; to apply the latter, we only need to construct a target that has a realistic ambiguity with a noun; the verb is gone already.
% However, in the context of real life texts, the grammar writer might actually mean the following:

% \begin{itemize}
% \item[] \begin{verbatim}
% REMOVE verb IF (-1 det) (0 noun) ;
% REMOVE noun IF (-1 det) (0 verb) ;
% \end{verbatim}
% \end{itemize}

% In that case, those rules are contradictory, and we would like to find it out.

>>>>

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
              & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
\# rules      & 59              & 279      & 1185     \\ \hline
\# readings   & 336             & 2191     & 2012    \\ \hline
\# conflicts  & 8 (1 + 5 + 2)   & 1        & 51    \\ \hline
\clock{} all rules       & 7s              & 3min 22s    & 3h 26min    \\ \hline
\clock{} last rule       & 4s              & 7s          & 3min 40s    \\ \hline
% This is with just 2000ish variables for Finnish, below is with 5700

@@

\end{tabular}
\caption{}
\label{table:res}
\end{table}

@@

The Spanish grammar had an erroneous set definition, where a word was
required to have almost all POS tags at the same time. After fixing
the set definition, our test did not reveal other problems.

The Dutch grammar had two kinds of errors: rule-internal due to nonexistent tags, and rule interaction. As for rule-internal conflicts, one was due to a misspelling in the list definition for personal pronouns, which rendered 5 rules ineffective. The other was about subreadings: the genitive \emph{s} is analysed as a subreading in the Apertium morphological analyser, but it appeared in 2 rules as the main reading.

There were two genuine conflicts with rule interaction. The first one is shown below:

\begin{itemize}
\item[] 
\begin{verbatim}REMOVE Adv IF (1 Noun) ;
REMOVE Adv IF (-1 Det) (0 Adj) (1 Noun) ;
\end{verbatim}
\end{itemize}

The first one is more straight-forward: two rules, which both remove an adverb, were in an order where the first has more broad condition: remove adverb if followed by a noun. In contrast, the second rule has a stricter requirement: only remove adverb if it is preceded by a determiner; the adverb itself is ambiguous with an adjective, and followed by a noun. The proble is that the first rule removes the adverb in all possible cases, and the second will not have any chance to act. If the rules were in the opposite order, then there would be no problem.

The second conflict, shown below, is more subtle.

\begin{itemize}
\item[] 
\begin{verbatim}SELECT DetPos IF (-1 (vbser pres p3 sg)) (0 ("zijn"i)) (1 Noun);
SELECT DetPosNotZijn IF (1 Noun);
\end{verbatim}
\end{itemize}

On the first glance it seems that there is no conflict: the first rule requires a Det Pos which is at the same time ``zijn'', as seen by the requirement \texttt{(0 ("zijn"i))}. However, this has a different semantics in the CG-3 formalism, as well as in SAT-CG. The rules match a situation where there is \emph{any} Det Pos and \emph{any} zijn, including the verb reading. The correct way would be

@@

### 2016-02-10 paper/lrec/implementation.tex
add stuff about new features

>>>>

There is another point to mention: the readings in grammar can be underspecified (e.g. `verb sg''), whereas the lexicon only gives us fully specified (``verb pres p2 sg'') readings. 
In addition, the sets in the grammar may include word forms or lemmas, but we remove those from the readings that we extract from the lexicon, in order to have 100s-1000s of variables instead of 100 000s.
A simple solution is to retain the lemmas and word forms only for those entries where it is specified in the grammar, and otherwise leave them out. For example, the grammar may contain the following list:

\begin{itemize}
 \item[] \texttt{LIST ZijnVerb = ("zijn" vbser pres pl) ;}
\end{itemize}

Clearly there is something special about the verb \emph{zijn}, not just any verb. Looking at the lexicon, we find \emph{zijn} in the following entries:

\begin{itemize}
 \item[] \texttt{zijn:zijn<det><pos><mfn><pl>}
 \item[] \texttt{zijn:zijn<det><pos><mfn><sg>}
 \item[] \texttt{zijn:zijn<vbser><inf>}
 \item[] \texttt{zijn:zijn<vbser><pres><pl>}
\end{itemize}

Thus we add special entries for these: in addition to the anonymous \texttt{<det><pos><mfn><pl>} reading, we add \texttt{<"zijn"><det><pos><mfn><pl>}.
